Data Specialization for a Staged Lambda Calculus

### 2015-02-05 paper/figures/splitting-overview.tex
wrote a new intro to splitting section
pared down implementation section

\caption{Summary of \lang\ evaluation.}
\label{fig:termSplitSummary}
\end{figure*}

@@

\end{abstrsyn}
\caption{Overview of Output Language.}
\label{fig:outputTermSummary}
\end{figure*}

@@

\put(28,66){\parbox[t]{44 \unitlength}{ 
	\textbf{Correctness Condition:} \\ 
	Splitting followed by evaluation (top and right arrows)
	should produce the same result as 
	evaluation followed by masking (left and bottom arrows),
	up to a term equivalence.
}}
\end{picture}

@@

\put(28,66){\parbox[t]{44 \unitlength}{ 
	\textbf{Correctness Condition:} \\ 
	Splitting followed by evaluation (top and right arrows)
	should produce the same result as 
	residualization (left and bottom arrow),
	up to a term equivalence.
}}
\end{picture}

@@

\end{abstrsyn}
\caption{Overview of Splitting.}
\label{fig:outputTermSummary}
\end{figure*}

>>>>

\caption{Summary of \lang\ evaluation.}
\label{fig:dynamicsSummary}
\end{figure*}

@@

\end{abstrsyn}
\caption{Overview of the output terms of splitting.}
\label{fig:outputTermSummary}
\end{figure*}

@@

\put(28,66){\parbox[t]{44 \unitlength}{ 
	\textbf{Correctness Condition:} \\ 
	Splitting followed by evaluation (top and right arrows)
	produce the same result as 
	evaluation followed by masking (left and bottom arrows),
	up to a term equivalence.
}}
\end{picture}

@@

\put(28,66){\parbox[t]{44 \unitlength}{ 
	\textbf{Correctness Condition:} \\ 
	Splitting followed by evaluation (top and right arrows)
	produce the same result as 
	residualization (left and bottom arrow),
	up to a term equivalence.
}}
\end{picture}

@@

\end{abstrsyn}
\caption{Overview of Splitting.}
\label{fig:splittingSummary}
\end{figure*}

### 2015-02-05 paper/implementation.tex
pared down implementation section

Applying the rules of \cref{sec:splitting} naively can yield inefficient split programs.
In many cases, the results can be improved by considering special cases in individual rules.
For instance, trivial subterm precomputations needn't be saved at tuples:
\[
\infer {\splittwo {(e_1,e_2)} A {p,l.(r_1,r_2)}}{\splittwo {e_1} A {(),\_.r_1} & \splittwo {e_2} A {p,l.r_2}}
\]

This sort of optimization does not affect the contract of splitting at all,
because the contract only requires that there be a precomputation that the residual can accept,
with no restrictions on the form of that precomputation.

The problem becomes more difficult when trying to optimize across function boundaries.
For example, consider the stage \bbtwo\ term,
\begin{lstlisting}
2`prev{
  1`letfun double (n : int) = n + n
  in next{`2`hold {`1`double 5`2`}-7`1`}`2`
}`
\end{lstlisting}
Using the rules above, this term splits into:
\begin{lstlisting}
1`letfun double (n : int) = (n+n, ())
in double 5`

2`l => 
letfun double (n : unit, l0 : unit) = ()
in (double (pi2 l); pi1 l)-7`
\end{lstlisting}

Observe that it's wasteful to run the stage \bbtwo\ version of \texttt{double}, since it always returns \texttt{()}. 
In this case, the wasted work is small, but if \texttt{double} were a recursive function,
then it's runtime may be unbounded.

There are two ways plausible ways to solve this problem:
either we could define an optimization step that comes 
after splitting and show that it tends to fix the inefficiencies that splitting creates; 
or we could give more structure to the input language so that splitting can generate more efficient code directly.

We choose to investigate the latter approach.
Specifically, we add new staging features to \lang\ which allow us to indicate that a piece of code is {\em strictly} 
stage \bbone---{\em i.e.} that it contains no $\next$ blocks.
At the type level, this manifests as a constructor called $\curr$.
So for instance, $\rmint \to \rmint$ is a stage \bbone\ function from one integer to another,
which might fail to terminate at stage \bbtwo, 
but $\curr (\rmint \to \rmint)$ is a stage \bbone\ function from one integer to another
which {\em cannot} have any content at stage \bbtwo.
Naturally, the well-formedness of $\curr A$ requires that $A$ contain no $\fut$s.

With these new features in place, we can now define the \texttt{double}
function so that it splits reasonably.
\begin{lstlisting}
2`prev{
  1`let mono{`3`double`1`} = mono{`3`fn n : int => n+n`1`}
  in next{`2`hold {`1`lift(mono{`3`double 5`1`})`2`}-7`1`}`2`
}`
\end{lstlisting}
This now splits into:
\begin{lstlisting}
1`let double = (fn n : int => n+n)
in (double 5,())`

2`l => in (pi1 l)-7`
\end{lstlisting}
\end{concretesyntax}

>>>>

Applying the rules of \cref{sec:splitting} naively can yield inefficient split programs.
In many cases, the results can be improved by considering special cases in individual rules.
For instance, trivial subterm precomputations needn't be saved at tuples:
\[
\infer {\splittwo {(e_1,e_2)} A {p,l.(r_1,r_2)}}{\splittwo {e_1} A {(),\_.r_1} & \splittwo {e_2} A {p,l.r_2}}
\]
Our implementation uses many optimizations like these.

\end{concretesyntax}
